HyperLogLog: Exponentially Bad in Adversarial Settings

### 2020-10-13 docs/HLL_exp_bad/main.tex
prepping for usenix submission
cutting to  13 pages

\title{HyperLogLog: Exponentially Bad in Adversarial Settings}
\author{Kenneth G. Paterson and Mathilde Raynal\\
\vspace{3mm}
Department of Computer Science, ETH Zurich\\
{\tt kenny.paterson@inf.ethz.ch, mraynal@student.ethz.ch}}

@@

\maketitle

%\IEEEtitleabstractindextext{%
\begin{abstract}
Computing the count of distinct elements in massive data sets is a common task but naive approaches are memory-expensive. The HyperLogLog (HLL) algorithm (Flajolet \emph{et al.}, 2007) estimates a data set's cardinality while using significantly less memory than a naive approach at the cost of some accuracy. This trade-off makes the HLL algorithm very attractive for a wide range of applications such as database management and network monitoring, where an exact count may not be needed. Recently, the HLL algorithm has started to be proposed for use in scenarios where the inputs may be adversarially generated, for example counting social network users or detection of network scanning attacks. This prompts an examination of the performance of the HLL algorithm in the face of adversarial inputs. We show that in such a setting, the HLL algorithm's estimate of cardinality can be exponentially bad: when an adversary has access to the internals of the HLL algorithm and has some flexibility in choosing what inputs will be recorded, it can manipulate the cardinality estimate to be exponentially smaller than the true cardinality. We study both the original HLL algorithm and a more modern version of it (Ertl, 2017) that is used in Redis. We present experimental results confirming our theoretical analysis. Finally, we consider attack prevention: we show to modify HLL in a simple way that provably prevents cardinality estimate manipulation attacks.
\end{abstract}
%}

@@

To complement our attack results, we provide a formal security analysis of HLL. We show that in the ``shadow device'' attack scenario of~\cite{hllvuln}, no security is possible even if the HLL is secretly keyed. On the other hand, we also show that in the most powerful attack setting introduced here, where the attacker has full access to the internals of the HLL (except for a secret key!), it is possible to provably secure HLL against cardinality manipulation by the simple expedient of replacing its internal hash function by a keyed, variable-input-length pseudo-random function (VIL-PRF). We describe a suitable instantiation of the required VIL-PRF using off-the-shelf cryptographic components.

@@

Both~\cite{cardestprivacy,hllvuln} propose mitigations using a salted sketch, either as a replacement for (but at the cost of losing mergeability) or in addition to (inducing a memory overhead) an unsalted sketch. We show in Section~\ref{sec:formal} that when a salt is used appropriately (i.e.\ as the secret key for a pseudo-random function that is used in place of hash function $h$), security against adversarial inputs can be achieved even in our strongest adversarial setting. 
%This might not be a convenient solution for some applications, they thus encourage further research.

@@

\begin{itemize}
\item[S1:] The adversary does not know the details of the target HLL implementation but can access a shadow copy of the HLL instance via its API. It can use the API to insert elements into this shadow sketch, get its cardinality estimate, and reset it to its native empty state. The adversary does not know the state of the targeted HLL sketch, meaning in particular that it does not have the list of elements previously inserted into it by other users. It can insert items into the target HLL sketch, but is otherwise ``blind''. 
\item[S2:] The adversary has access to the details of the HLL implementation, i.e.\ it knows the number of buckets $m=2^n$ and the hash function $h$ in use. It can insert items into the HLL sketch, but is otherwise blind. In particular, an S2 adversary does not have access to the values of the HLL cardinality estimate at any point in its attack. This kind of attack scenario is appropriate in the context of the port scanning attack of~\cite{portscanhll}.
\item[S3:] The adversary has access to the details of the HLL implementation, i.e.\ it knows the number of buckets $m=2^n$ and the hash function $h$ in use. Additionally, it can access and interact with the sketch via an API provided by the sketch owner, allowing it to insert items into the HLL sketch and ask for the HLL's cardinality estimate at any point in time. However, it does not know the individual bucket contents.
\item[S4:] The adversary has direct access to the HLL in use and all its internals, i.e., the number of buckets, the hash function $h$, and also the contents of each bucket in the HLL sketch \emph{at one specific point in time}, i.e.\ it is given a \emph{snapshot} of the sketch. It can access and interact with the sketch via an API provided by the sketch owner, as per scenario S3. Note that, using these capabilities, an S4 adversary can compute a lower bound on the HLL cardinality estimate based on its initial snapshot and whatever it causes to be inserted into the HLL sketch during its attack (this is irrespective of whether its attack proceeds with input from other users being added to the HLL sketch in parallel with its attack or not; if it knows that there is no other input, then it can compute the \emph{exact} value of the HLL cardinality estimate at every point in its attack instead of a lower bound).
\end{itemize}

@@

\begin{table}[h]
\centering
\caption{Adversarial capabilities in scenarios S1, S2, S3 and S4.}
\medskip
     \makebox[\linewidth]{
\begin{tabular}{| m{11em} | m{2em} | m{2em} | m{2em} | m{2em} |}
    \hline
    {Scenario} 							&  {S1} & {S2} & {S3} & {S4} \\ \hline
   	{API for shadow device} 		& \cmark & \xmark & \xmark & \xmark \\ \hline
    {Insertion of  items} 				& \cmark & \cmark & \cmark & \cmark \\ \hline
    {$h$ and $m$ known}			& \xmark & \cmark & \cmark & \cmark \\ \hline
    {Get cardinality estimate} 		& \xmark & \xmark & \cmark & \cmark \\ \hline
    {Get snapshot} 						& \xmark & \xmark & \xmark & \cmark \\ \hline
\end{tabular}}
\label{tab:scenarios}
\end{table}

@@

Scenario S1 is rather distinctive in that it is the only one in which the adversary has access via an API to a shadow device (and no other capabilities). We do not consider scenario S1 typical of HLL deployments. In a networking application, the authors of~\cite{hllvuln} argue that the assumptions of S1 could be fulfilled if the adversary learns which machine is used to do the monitoring, buys the same, and uses it as the shadow device. It is possible that the provider uses a different salt or key in the hash per device. This would make any pre-computation on the items using the shadow device irrelevant, and considerably weaken the attack presented in Section 3.2 of~\cite{hllvuln}. Hence we assume for the rest of this work that, in scenario S1, shadow and targeted devices have the same internals and security parameters (if there are any). If the HLL device receives inputs from several users, the attack from Section 3.2 of~\cite{hllvuln} is then the easiest to carry out. It is described in~\cite{hllvuln} as follows: "after each insertion [in the shadow device], elements that do not increase the cardinality estimate are retained". The idea is to iterate through the process of trial insertion several times to refine the list of items, before inserting what remains into the targeted device. We explore in this work scenario S1 in the special case where inputs come only from the adversary, and propose an attack that provides significantly better results than the one presented in~\cite{hllvuln}.

@@

However, this procedure does not work as simply as just described because of numerical issues that arise in the computation of the cardinality estimate. Specifically, increasing the value held in some bucket from $M[j]$ to $M[j]+1$ \emph{decreases} the denominator in the raw cardinality estimate $\sum_{j=1}^m2^{-M[j]}$ by an amount $2^{-M[j]-1}$. This does lead to a change in the raw cardinality estimate $E = \alpha_mm^2\cdot(\sum_{j=1}^m2^{-M[j]})^{-1}$. However, in practice, the raw cardinality estimate is rounded before being presented to the adversary, so this change may not actually be detectable. We have done experiments which shows this to be the case for parameters arising in practice.

@@

As a second example, setting $B = 1$, we can expect to insert $2^{t-n}$ items with a final cardinality estimate equal to $m \cdot \ln (1+\frac{1}{m-1})$, which is rounded to $1$ by the HLL algorithm (for $m$ of practical interest, e.g.\ $m=256$). Thus we keep the HLL cardinality estimate to a constant (1!) whilst being able to insert exponentially many items, provided there is sufficient flexibility in the choice of input set ${\cal X}$.

@@

Now we expect to reject in total $(1 - \frac{B}{m})\cdot 2^t$ items, and each one needs an expected number of trial insertions that is bounded by $m/(m - B_{\max} + B)$. (In fact, we can model the number of trial insertions needed as an independent geometric distribution with parameter $q$ for each such $x$.) Meanwhile, each of the $\frac{B}{m}\cdot 2^t$ items in the final insertion list ${\cal L}^*$ for the target HLL sketch is trial inserted into the shadow HLL sketch on every pass.

@@

Now we turn to the case where the HLL sketch under attack is not initially empty, but instead contains some honestly inserted items. As usual, the adversary's goal is to insert as many items as possible whilst increasing the cardinality estimate as little as possible.

We focus here on scenarios S2 and S4, and make brief remarks on S1.

We assume that the target HLL sketch in S2 and S4 has already had sufficiently many items inserted that every one of the $m=2^n$ buckets has counter value $M[j]$ with $M[j] \ge 1$. This is highly likely to be the case (with probability at least $1-1/m^2$) as long as at least $m\ln m$ items have already been inserted into the target HLL sketch. This follows from the standard analysis of the coupon collector problem under our usual assumption that $h$ behaves as a random function.
In S4, the adversary can use its snapshot to detect if this is the case or not, and behave differently if not (details for this case follow below).

With $m \ln m$ honest insertions, it is also highly likely that the small range correction (Hit Counting algorithm) will not be used, and the target HLL's cardinality estimate will be computed using the raw estimate:
\[
E = \alpha_mm^2\cdot(\sum_{j=1}^m2^{-M[j]})^{-1}.
\]

We assume the adversary has (or is given) some set of items ${\cal X}$ that it wishes to insert; as usual, we write $|{\cal X}| = 2^t$.

\subsubsection{Initially Non-empty HLL Sketch, Scenario S2:}\label{sec:nonemptyHLLS2}
The adversary knows $h$ so it selects items $x$ from ${\cal X}$ for which the $\ell$ rightmost bits of $h(x)$ start with a ``1'' bit. In this case, since the relevant bucket $j$ already has $M[j] \ge 1$, the bucket's counter is not updated, and the cardinality estimate of the HLL sketch remains unchanged.

The adversary then expects to be able to insert half of the items from ${\cal X}$ into the HLL sketch, without increasing the HLL cardinality estimate at all.

@@

\subsubsection{Initially Non-empty HLL Sketch, Scenario S4:}\label{sec:nonemptyHLLS4}
The adversary can adopt the same strategy as for S2, but make relax the sampling requirements with the additional information available in this scenario (namely, the count values $M[j]$ per bucket in the HLL sketch). Specifically, it can allow some leading zero bits in the $\ell$ rightmost bits of $h(x)$ so long as this does not increase the current bucket estimate. In other words, if $h(x)$ is mapped to the $j$-th bucket containing $M[j] \ge 1$, then, instead of the strict requirement of having a ``1'' bit in the first position in the $\ell$ rightmost bits of $h(x)$, the adversary is satisfied when there is a ``1'' bit in any of the first $M[j]$ positions in this substring. Under these conditions, the estimates in the buckets are never updated, and there is no effect on the cardinality estimate made by HLL.

We now compute the expected number of elements $x$ from ${\cal X}$ such that $h(x)$ meets the attack's requirements. We assume that $M[j] \ge 1$ for all buckets $j$. Assume $h(x)$ is mapped to bucket $j$. Inserting $x$ does not increase the cardinality estimate if the rightmost $\ell$ bits of $h(x)$ have $M[j]-1$ or fewer leading zeroes; this probability is equal to $1-\frac{1}{2^{M[j]}}$ assuming $h$ behaves like a random function. Averaging over all $m$ buckets (and using the fact that the bucket choice is uniformly random since the bits of $h$ are independent and uniformly random), the probability that input $x$ meets the attack's requirements is given by:
\begin{eqnarray*}
\sum_{j=1}^{m}\frac{1}{m} \cdot (1-\frac{1}{2^{M[j]}}) & = & 1- \frac{1}{m}\cdot \sum_{j=1}^{m} {(2^{M[j]})}^{-1} \\
& = &1 - H_M^{-1} 
\end{eqnarray*}
where $H_M$ is the harmonic mean of the counts $2^{M[1]}, 2^{M[2]}, \ldots, 2^{M[m]}$.
So, given ${\cal X}$ containing $2^t$ items, the adversary can expect to insert $(1 - H_M^{-1}) \cdot 2^t$ items from ${\cal X}$ without increasing the HLL cardinality estimate at all.

If some buckets are actually empty (and in S4 the adversary knows which ones using its snapshot), then the above analysis can be reused in order to avoid touching those buckets altogether when inserting elements from ${\cal X}$. The above analysis does not change at all because the term $\frac{1}{m} \cdot (1-\frac{1}{2^{M[j]}})$ already signs a probability of 0 to inserting $x$ in bucket $j$ if $M[j] = 0$. Hence we still expect to insert $(1 - H_M^{-1}) \cdot 2^t$ items from ${\cal X}$ without increasing the HLL cardinality estimate at all. Of course, the adversary can relax its attack and allow some of the empty buckets to be hit as well. This would increase the insertion rate at the cost of an increase in the final cardinality estimate.

In this setting, where the target HLL sketch is initially non-empty, we see that an adversary who has sufficient flexibility in its inputs (as represented by the size $2^t$ of the set ${\cal X}$) can expect to insert exponentially many (in $t$) values into the HLL sketch without increasing the cardinality estimate (or whilst increasing the cardinality estimate slightly in the case where the HLL sketch has some empty buckets in S2). The result is largely independent of the HLL parameters, but does depend on the extent to which the HLL sketch is already filled by other users, via a factor $1-H_M^{-1}$: the more full the HLL sketch is, the easier it is to insert new items without them increasing the cardinality estimate.

@@

Finally, we briefly consider scenario S1 with an initially non-empty target HLL sketch. Since the target sketch has an unknown status for the adversary, our previous strategy for S1 of keeping many buckets empty does not work. One can attempt to use the shadow device to identify items that would result in small entries $M[j]$ (for unknown indices $j$) since such items are unlikely to increase the cardinality estimate of the target device by much.  This requires first filling the shadow device with sufficiently many values to escape from the ``small range correction'' phase. Then one could enter random values $x$ into the shadow device and check for cardinality increases. However this runs into the rounding issue described in Section~\ref{sec:scenarios-discussion}. Still, one could retain those items which do lead to a detectable change in cardinality and reject the others (and also estimate the value of $M[j]$ for the unknown bucket $j$ from the magnitude of the change, retaining only those for which $M[j]$ is small). But here the lack of a cardinality change could result either from rounding issues or because insertion of $x$ into the shadow device really did lead to no increase in cardinality. This indicates that  ``good'' items would be rejected unnecessarily. This could be addressed by using multiple resets, filling with random items, and then multiple insertions per item to try to estimate the true value of $M[j]$ for each item. The details of such an attack require further analysis which we do not pursue here, since we consider S1 to be the least realistic attack scenario.

@@

Ertl~\cite{hllnew} proposed using the following cardinality estimator in place of the raw estimator in Figure~\ref{fig:hll}:
\begin{equation}\label{eqn:ertl}
\hat{\lambda}  = \frac{\alpha_\infty m^2}{m\sigma(C_0/m) + \sum_{k=1}^{\ell}C_k 2^{-k} + m \tau(1-C_{\ell+1}/m)2^{-\ell}} \, .
\end{equation}
Here, in the denominator, $C_k$ denotes the number of buckets containing value $k$ in the HLL filter, so $C_0$ is in particular the number of empty buckets; $\sigma(\cdot)$ and $\tau(\cdot)$ are specific functions defined as:
\[
\sigma(x) :=  x + \sum_{k=1}^{\infty} x^{2^k} 2^{k-1}
\]
and
\[
\tau(x) :=  \frac{1}{2} \left( -x + \sum_{k=1}^{\infty} x^{2^{-k}} 2^{-k} \right).
\]
Finally,  $\alpha_\infty  = 0.72134752\ldots$ is a constant. (Note that we have translated from the original notation in~\cite{hllnew} to ours.)

@@

Note that when evaluating (\ref{eqn:ertl}), functions $\sigma$ and $\tau$ are only called on inputs in the range $[0,1]$. Moreover, in this range, we have $\sigma(x) \ge x$. Also it is highly likely that $C_{\ell+1} = 0$ (since it is very unlikely that a 50-bit all-0 string will result from the application of the Redis HLL hash function, and in any case in scenarios S2 and S4, we can always arrange for this to be avoided). When $C_{\ell+1} = 0$, $\tau$ is evaluated at 1, where its value is zero. Combining this information, we see that, provided $C_{\ell+1} = 0$, and letting $D$ denote the denominator in equation~(\ref{eqn:ertl}), we have:
\[
D \ge m\sigma(C_0/m) \ge C_0
\]
and so
\[
\hat{\lambda}  \leq \frac{\alpha_\infty m^2}{C_0}
\]
where, recall, $C_0$ is the number of empty buckets.

@@

The requirement on $C_0 \geq {\epsilon}m$ is easily met in scenarios S2 and S4 with an initially empty HLL filter, through the trick of sampling inputs $x \in {\cal X}$ so that only a fraction $1-\epsilon$ of the buckets are ever hit. We can expect to insert a fraction $1-\epsilon-\epsilon2^{-\ell}$ of the elements of  ${\cal X}$ whilst keeping the cardinality estimate $\hat{\lambda}$ below  $\alpha_\infty \epsilon^{-1} m$. (The term $\epsilon2^{-\ell}$ arises from the requirement to keep $C_{\ell+1} =0$ in the attack; recall that $\ell=50$ in Redis, so the effect of this term is very small.)

>>>>

\title{HyperLogLog: Exponentially Bad in Adversarial Settings}
%\author{Kenneth G. Paterson and Mathilde Raynal\\
%\vspace{3mm}
%Department of Computer Science, ETH Zurich\\
%{\tt kenny.paterson@inf.ethz.ch, mraynal@student.ethz.ch}}

@@

\maketitle
%\IEEEtitleabstractindextext{%

\begin{abstract}
Computing the count of distinct elements in massive data sets is a common task but naive approaches are memory-expensive. The HyperLogLog (HLL) algorithm (Flajolet \emph{et al.}, 2007) estimates a data set's cardinality while using significantly less memory than a naive approach at the cost of some accuracy. This trade-off makes the HLL algorithm very attractive for a wide range of applications such as database management and network monitoring, where an exact count may not be needed. Recently, the HLL algorithm has started to be proposed for use in scenarios where the inputs may be adversarially generated, for example counting social network users or detection of network scanning attacks. This prompts an examination of the performance of the HLL algorithm in the face of adversarial inputs. We show that in such a setting, the HLL algorithm's estimate of cardinality can be exponentially bad: when an adversary has access to the internals of the HLL algorithm and has some flexibility in choosing what inputs will be recorded, it can manipulate the cardinality estimate to be exponentially smaller than the true cardinality. We study both the original HLL algorithm and a more modern version of it (Ertl, 2017) that is used in Redis. We present experimental results confirming our theoretical analysis. Finally, we consider attack prevention: we show to modify HLL in a simple way that provably prevents cardinality estimate manipulation attacks.
\end{abstract}
%}

@@

To complement our attack results, we provide a formal security analysis of HLL. We show that in the ``shadow device'' attack scenario of~\cite{hllvuln}, no security is possible even if the HLL is secretly keyed. On the other hand, we also show that in the most powerful attack setting introduced here, where the attacker has full access to the internals of the HLL algorithm (except for a secret key!), it is possible to provably secure HLL against cardinality manipulation by the simple expedient of replacing its internal hash function by a keyed, variable-input-length pseudo-random function (VIL-PRF). We describe a suitable instantiation of the required VIL-PRF using off-the-shelf cryptographic components.

@@

Both~\cite{cardestprivacy,hllvuln} propose mitigations using a salted HLL sketch, either as a replacement for (but at the cost of losing mergeability) or in addition to (inducing a memory overhead) an unsalted HLL. We show in Section~\ref{sec:formal} that when a salt is used appropriately (i.e.\ as the secret key for a pseudo-random function that is used in place of hash function $h$), security against adversarial inputs can be achieved even in our strongest adversarial setting. 
%This might not be a convenient solution for some applications, they thus encourage further research.

@@

\begin{itemize}
\item[S1:] The adversary does not know the details of the target HLL implementation but can access a shadow copy of the HLL instance via its API. It can use the API to insert elements into this shadow sketch, get its cardinality estimate, and reset it to its native empty state. The adversary does not know the state of the targeted HLL sketch, meaning in particular that it does not have the list of elements previously inserted into it by other users. It can insert items into the target HLL sketch, but is otherwise ``blind''. 
\item[S2:] The adversary has access to the details of the HLL implementation, i.e.\ it knows the number of buckets $m=2^n$ and the hash function $h$ in use. It can insert items into the HLL sketch, but is otherwise blind. In particular, an S2 adversary does not have access to the values of the HLL cardinality estimate at any point in its attack. This kind of attack scenario is appropriate in the context of the port scanning attack of~\cite{portscanhll}.
\item[S3:] The adversary has access to the details of the HLL implementation, i.e.\ it knows the number of buckets $m=2^n$ and the hash function $h$ in use. Additionally, it can access and interact with the sketch via an API provided by the sketch owner, allowing it to insert items into the HLL sketch and ask for the HLL's cardinality estimate at any point in time. However, it does not know the individual bucket contents.
\item[S4:] The adversary has direct access to the HLL in use and all its internals, i.e., the number of buckets, the hash function $h$, and also the contents of each bucket in the HLL sketch \emph{at one specific point in time}, i.e.\ it is given a \emph{snapshot} of the sketch. It can access and interact with the sketch via an API provided by the sketch owner, as per scenario S3. 
%Note that, using these capabilities, an S4 adversary can compute a lower bound on the HLL cardinality estimate based on its initial snapshot and whatever it causes to be inserted into the HLL sketch during its attack (this is irrespective of whether its attack proceeds with input from other users being added to the HLL sketch in parallel with its attack or not; if it knows that there is no other input, then it can compute the \emph{exact} value of the HLL cardinality estimate at every point in its attack instead of a lower bound).
\end{itemize}

@@

\begin{table}[tb!]
\small
\centering
\caption{Adversarial capabilities in scenarios S1, S2, S3 and S4.}
\medskip
     \makebox[\linewidth]{
\begin{tabular}{| m{11em} | m{2em} | m{2em} | m{2em} | m{2em} |}
    \hline
    {Scenario} 							&  {S1} & {S2} & {S3} & {S4} \\ \hline
   	{API for shadow device} 		& \cmark & \xmark & \xmark & \xmark \\ \hline
    {Insertion of  items} 				& \cmark & \cmark & \cmark & \cmark \\ \hline
    {$h$ and $m$ known}			& \xmark & \cmark & \cmark & \cmark \\ \hline
    {Get cardinality estimate} 		& \xmark & \xmark & \cmark & \cmark \\ \hline
    {Get snapshot} 						& \xmark & \xmark & \xmark & \cmark \\ \hline
\end{tabular}}
\label{tab:scenarios}
\end{table}

@@

Scenario S1 is distinctive in that it is the only one in which the adversary has access via an API to a shadow device (and no other capabilities). We do not consider scenario S1 typical of HLL deployments. In a networking application, the authors of~\cite{hllvuln} argue that the assumptions of S1 could be fulfilled if the adversary learns which machine is used to do the monitoring, buys the same, and uses it as the shadow device. It is possible that the provider uses a different salt or key in the hash per device. This would make any pre-computation on the items using the shadow device irrelevant, and considerably weaken the attack presented in Section 3.2 of~\cite{hllvuln}. Hence we assume for the rest of this work that, in scenario S1, shadow and targeted devices have the same internals and security parameters (if there are any). 
%If the HLL device receives inputs from several users, the attack from Section 3.2 of~\cite{hllvuln} is then the easiest to carry out. It is described in~\cite{hllvuln} as follows: "after each insertion [in the shadow device], elements that do not increase the cardinality estimate are retained". The idea is to iterate through the process of trial insertion several times to refine the list of items, before inserting what remains into the targeted device. We explore in this work scenario S1 in the special case where inputs come only from the adversary, and propose an attack that provides significantly better results than the one presented in~\cite{hllvuln}.

@@

However, this procedure does not work as simply as just described because of numerical issues that arise in the computation of the cardinality estimate. Specifically, increasing the value held in some bucket from $M[j]$ to $M[j]+1$ \emph{decreases} the denominator in the raw cardinality estimate $\sum_{j=1}^m2^{-M[j]}$ by an amount $2^{-M[j]-1}$. This does lead to a change in the raw cardinality estimate $E = \alpha_mm^2\cdot(\sum_{j=1}^m2^{-M[j]})^{-1}$. However, in practice, the raw cardinality estimate is rounded before being presented to the adversary, so this change may not actually be detectable, especially when $M[j]$ becomes large. We have done experiments which shows this to be the case for parameters arising in practice.

@@

As a second example, setting $B = 1$, we can expect to insert $2^{t-n}$ items with a final cardinality estimate equal to $m \cdot \ln (1+\frac{1}{m-1})$, which is rounded to $1$ by the HLL algorithm (for $m$ of practical interest, e.g.\ $m=256$). Thus we keep the HLL cardinality estimate to a constant (1!) whilst being able to insert exponentially many items, provided there is sufficient flexibility in the input set ${\cal X}$.

@@

Now we expect to reject in total $(1 - \frac{B}{m})\cdot 2^t$ items, and each one needs an expected number of trial insertions that is bounded by $m/(m - B_{\max} + B)$. 
%(In fact, we can model the number of trial insertions needed as an independent geometric distribution with parameter $q$ for each such $x$.) 
Meanwhile, each of the $\frac{B}{m}\cdot 2^t$ items in the final insertion list ${\cal L}^*$ for the target HLL sketch is trial inserted into the shadow HLL sketch on every pass.

@@

Now we turn to the case where the HLL under attack is not initially empty, but instead contains some honestly inserted items. As usual, the adversary's goal is to insert as many items as possible whilst increasing the cardinality estimate as little as possible. We focus on scenarios S2 and S4, and make brief remarks on S1.

We assume that the target HLL in S2 and S4 has already had sufficiently many items inserted that every one of the $m=2^n$ buckets has counter value $M[j]$ with $M[j] \ge 1$. This is highly likely to be the case (with probability at least $1-1/m^2$) as long as at least $m\ln m$ items have already been inserted into the target HLL. This follows from the standard analysis of the coupon collector problem under our usual assumption that $h$ behaves as a random function.
In S4, the adversary can use its snapshot to detect if this is the case or not, and behave differently if not (details for this case follow below).

With $m \ln m$ honest insertions, it is also highly likely that the small range correction (Hit Counting algorithm) will not be used, and the target HLL's cardinality estimate will be computed using the raw estimate.
%\[
%E = \alpha_mm^2\cdot(\sum_{j=1}^m2^{-M[j]})^{-1}.
%\]

%We assume the adversary has (or is given) some set of items ${\cal X}$ that it wishes to insert; as usual, we write $|{\cal X}| = 2^t$.

\subsubsection{Initially Non-empty HLL Sketch, Scenario S2:}\label{sec:nonemptyHLLS2}
The adversary knows $h$ so it selects items $x$ from ${\cal X}$ for which the $\ell$ rightmost bits of $h(x)$ start with a ``1'' bit. In this case, since the relevant bucket $j$ already has $M[j] \ge 1$, the bucket's counter is not updated, and the cardinality estimate of the HLL remains unchanged.

The adversary then expects to be able to insert half of the items from ${\cal X}$ into the HLL, without increasing the HLL cardinality estimate at all.

@@

\subsubsection{Initially Non-empty HLL Sketch, Scenario S4:}\label{sec:nonemptyHLLS4}
The adversary can adopt the same strategy as for S2, but make relax the sampling requirements with the additional information available in this scenario (namely, the count values $M[j]$ per bucket in the HLL sketch). Specifically, it can allow some leading zero bits in the $\ell$ rightmost bits of $h(x)$ so long as this does not increase the current bucket estimate. In other words, if $h(x)$ is mapped to the $j$-th bucket containing $M[j] \ge 0$, then, instead of the strict requirement of having a ``1'' bit in the first position in the $\ell$ rightmost bits of $h(x)$, the adversary is satisfied when there is a ``1'' bit in any of the first $M[j]$ positions in this substring. (A value $M[j]=0$ then never permits insertions, meaning an empty bucket stays empty.) Under these conditions, the estimates in the buckets are never updated, and there is no effect on the cardinality estimate made by HLL.

We now compute the expected number of elements $x$ from ${\cal X}$ such that $h(x)$ meets the attack's requirements.  %We assume that $M[j] \ge 0$ for all buckets $j$. 
Assume $h(x)$ is mapped to bucket $j$. Inserting $x$ does not increase the cardinality estimate if the rightmost $\ell$ bits of $h(x)$ have $M[j]-1$ or fewer leading zeroes; this probability is equal to $1-\frac{1}{2^{M[j]}}$ assuming $h$ behaves like a random function. Averaging over all $m$ buckets (and using the fact that the bucket choice is uniformly random since the bits of $h$ are independent and uniformly random), the probability that input $x$ meets the attack's requirements is given by:
\begin{eqnarray*}
\sum_{j=1}^{m}\frac{1}{m} \cdot (1-\frac{1}{2^{M[j]}}) & = & 1- \frac{1}{m}\cdot \sum_{j=1}^{m} {(2^{M[j]})}^{-1} \\
& = &1 - H_M^{-1} 
\end{eqnarray*}
where $H_M$ is the harmonic mean of the counts $2^{M[1]}, 2^{M[2]}, \ldots, 2^{M[m]}$.
So, given ${\cal X}$ containing $2^t$ items, the adversary can expect to insert $(1 - H_M^{-1}) \cdot 2^t$ items from ${\cal X}$ without increasing the HLL cardinality estimate at all.

%If some buckets are actually empty (and in S4 the adversary knows which ones using its snapshot), then the above analysis can be reused in order to avoid touching those buckets altogether when inserting elements from ${\cal X}$. The above analysis does not change at all because the term $\frac{1}{m} \cdot (1-\frac{1}{2^{M[j]}})$ already signs a probability of 0 to inserting $x$ in bucket $j$ if $M[j] = 0$. Hence we still expect to insert $(1 - H_M^{-1}) \cdot 2^t$ items from ${\cal X}$ without increasing the HLL cardinality estimate at all. Of course, the adversary can relax its attack and allow some of the empty buckets to be hit as well. This would increase the insertion rate at the cost of an increase in the final cardinality estimate.

In this setting, where the target HLL sketch is initially non-empty, we see that an adversary who has sufficient flexibility in its inputs (as represented by the size $2^t$ of the set ${\cal X}$) can expect to insert exponentially many (in $t$) values into the HLL sketch without increasing the cardinality estimate (or whilst increasing the cardinality estimate slightly in the case where the HLL sketch has some empty buckets in S2). The result is largely independent of the HLL parameters. The more full is the HLL sketch, the larger is $1-H_M^{-1}$ and the easier it is to insert new items without them increasing the cardinality estimate.

@@

We briefly sketch an attack in this setting in Appendix~\ref{sec:non-empty-HLL-sketch-S1}.

%Finally, we briefly consider scenario S1 with an initially non-empty target HLL sketch. Since the target sketch has an unknown status for the adversary, our previous strategy for S1 of keeping many buckets empty does not work. One can attempt to use the shadow device to identify items that would result in small entries $M[j]$ (for unknown indices $j$) since such items are unlikely to increase the cardinality estimate of the target device by much.  This requires first filling the shadow device with sufficiently many values to escape from the ``small range correction'' phase. Then one could enter random values $x$ into the shadow device and check for cardinality increases. However this runs into the rounding issue described in Section~\ref{sec:scenarios-discussion}. Still, one could retain those items which do lead to a detectable change in cardinality and reject the others (and also estimate the value of $M[j]$ for the unknown bucket $j$ from the magnitude of the change, retaining only those for which $M[j]$ is small). But here the lack of a cardinality change could result either from rounding issues or because insertion of $x$ into the shadow device really did lead to no increase in cardinality. This indicates that  ``good'' items would be rejected unnecessarily. This could be addressed by using multiple resets, filling with random items, and then multiple insertions per item to try to estimate the true value of $M[j]$ for each item. The details of such an attack require further analysis which we do not pursue here, since we consider S1 to be the least realistic attack scenario.

@@

Ertl~\cite{hllnew} proposed using the following cardinality estimator in place of the raw estimator in Figure~\ref{fig:hll}:
\begin{equation}\label{eqn:ertl}
\hat{\lambda}  = \frac{\alpha_\infty m^2}{m\sigma(C_0/m) + \sum_{k=1}^{\ell}C_k 2^{-k} + m \tau(1-C_{\ell+1}/m)2^{-\ell}} \, .
\end{equation}
Here, in the denominator, $C_k$ denotes the number of buckets containing value $k$ in the HLL sketch, so $C_0$ is in particular the number of empty buckets; $\sigma(\cdot)$ and $\tau(\cdot)$ are specific functions defined as:
\[
\sigma(x) :=  x + \sum_{k=1}^{\infty} x^{2^k} 2^{k-1}
\]
and
\[
\tau(x) :=  \frac{1}{2} \left( -x + \sum_{k=1}^{\infty} x^{2^{-k}} 2^{-k} \right).
\]
Finally,  $\alpha_\infty  = 0.72134752\ldots$ is a constant. (Note that we have translated from the original notation in~\cite{hllnew} to ours.)

@@

Note that when evaluating (\ref{eqn:ertl}), functions $\sigma$ and $\tau$ are only called on inputs in the range $[0,1]$. Moreover, in this range, we have $\sigma(x) \ge x$. It is also highly likely that $C_{\ell+1} = 0$ (since it is very unlikely that a 50-bit all-0 string will result from the application of the Redis HLL hash function, and in scenarios S2 and S4, we can always arrange for this to be avoided). When $C_{\ell+1} = 0$, $\tau$ is evaluated at 1, where its value is zero. Combining this information, we see that, provided $C_{\ell+1} = 0$, and letting $D$ denote the denominator in equation~(\ref{eqn:ertl}), we have:
\[
D \ge m\sigma(C_0/m) \ge C_0
\]
and so
\[
\hat{\lambda}  \leq \frac{\alpha_\infty m^2}{C_0}
\]
where, recall, $C_0$ is the number of empty buckets.

@@

The requirement on $C_0 \geq {\epsilon}m$ is easily met in scenarios S2 and S4 with an initially empty HLL sketch, through the trick of sampling inputs $x \in {\cal X}$ so that only a fraction $1-\epsilon$ of the buckets are ever hit. We can expect to insert a fraction $1-\epsilon-\epsilon2^{-\ell}$ of the elements of  ${\cal X}$ whilst keeping the cardinality estimate $\hat{\lambda}$ below  $\alpha_\infty \epsilon^{-1} m$. (The term $\epsilon2^{-\ell}$ arises from the requirement to keep $C_{\ell+1} =0$ in the attack; recall that $\ell=50$ in Redis, so the effect of this term is very small.)

In scenarios S2 and S4 with an initially non-empty HLL sketch, we assume as for our earlier analysis of the standard HLL algorithm that all buckets have already been hit (this will be the case with high probability as soon as $m \log m$ items have been honestly inserted). Hence $C_0 = 0$ and $\sigma$ is evaluated at 0 in (\ref{eqn:ertl}), where its value is 0. This means that the estimate $\hat{\lambda}$ is computed using a denominator of the form:
\[
\sum_{k=1}^{\ell}C_k 2^{-k} + m \tau(1-C_{\ell+1}/m)2^{-\ell}.
\]
Now we can proceed exactly as in Sections~\ref{sec:nonemptyHLLS2} and~~\ref{sec:nonemptyHLLS4}, inserting values $x$ whose hashes have rightmost $\ell$ bits beginning with a ``1'' bit (and with a more relaxed condition in S4). These insertions do not alter the values of the $C_k$, and hence $\hat{\lambda}$ is unaffected. In S2, we expect to insert half the items in ${\cal X}$ without increasing the cardinality estimate $\hat{\lambda}$ at all. In S4, the analysis is the same as in Section~\label{sec:nonemptyHLLS4}.

This completes our analysis of Redis. The key takeaway is that, in scenarios S2 and S4, an adversary can expect to insert a constant fraction of ${\cal X}$ into a non-empty Redis HLL sketch whilst not increasing its cardinality estimate at all (and whilst keeping the estimated cardinality linear in $m$ in the case where the sketch is initially empty).
